\section{Organization of this report}

The steady-state Boltzmann transport equation with a fixed source term will be stated in Chapter 2 in both continuous 
energy and multigroup form. Following \cite{DautrayLions}, we will review conditions for the existence of a unique 
non-negative solution of the equation.

Main contribution of this preliminary dissertation report to the current state of neutron transport knowledge is 
contained in Chapter 3 where the \textit{Maxwell-Cartesian surface spherical harmonic functions} are used to derive an
alternative to the classical set of $\PN$ equations (the resulting set is called here ``$\MCPN$"). Despite its
favorable properties with respect to the traditional basis of tesseral spherical harmonics, these functions have not
been used in a systematic way for neutron transport problems before (the few related works will be dissected in Chap. 3
and main differences from the present work underlined). A very close relationship between the form of the $\MCPN$ 
equations and that of the 1D $\PN$ equations is then noted, possibly bringing new light into the question of validity
of the $\SPN$ approximation. The structure of the $\MCPN$ set also allows manipulations which could lead to other 
advantageous forms of the $\PN$ equations that were not thoroughly explored before -- one of them is outlined at the end
of the chapter.

Numerical methods for solving either the $\MCPN$ set or its derived forms (including the $\SPN$ equations) will be 
formulated and analyzed in the final dissertation thesis. Outline of this future work concludes this report.
