% Thesis Acknowledgements ------------------------------------------------


%\begin{acknowledgementslong} %uncommenting this line, gives a different acknowledgements heading
\begin{acknowledgements}      %this creates the heading for the acknowlegments

I would like to thank my supervisor Marek Brandner, for all his support and guidance throughout the course of 
my Ph.D. studies. My thanks also belong to all members of our TACR-PAMG team, especially to the ``Himalaya
Expedition group'' Roman, Hanka and Zby{\v n}{\' a}k, who influenced my thinking in so many ways both direct and
indirect. In particular, I would like to thank Roman for explaining to me many of his great ideas (not only about
homebrewing). Having said so, I cannot fail to mention how Hanka took care of us when we were so deeply immersed in discussing those ideas, for which I (and Roman certainly too)
am so grateful.

I would also like to thank all the great people whom I had an honor to work with during my Ph.D. studies and were
not mentioned above. Namely to Luk{\' a}{\v s} Korous, with whom I spent so many productive days working on Hermes,
Pavel {\v S}ol{\' i}n for inviting me to Nevada to work on Hermes and him and his lovely wife D{\' a}{\v s}a for
letting me stay in their house during that visit, and Vyacheslav G. Zimin for
showing me and Roman the meaning of Russian hospitality during our research visit to the International Science \&
Technology Center in Moscow. Special thanks go to Ryan G. McClarren from the Texas A\&M University in College Station who made it
possible for me to meet the transport theory experts (including himself) and made each of my visits a smooth and
enjoyable experience.

Last but certainly not least, I would like to express my deep gratitude to my family, who have always supported me in
any way they could.

\end{acknowledgements}
%\end{acknowledgmentslong}

% ------------------------------------------------------------------------


